---
title: "Case Study: CELLxGENE Ontologies"
author:
- name: Martin Morgan
  affiliation: "Roswell Park Comprehensive Cancer Center, Buffalo, NY, US"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Case Study: CELLxGENE Ontologies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r vignette-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
## silence progress bar in the vignette
options(cli.progress_show_after = 900)
```

This case study illustrates use of OLSr to facilitate data set
discovery in the [CELLxGENE data portal][cxgdp]. The CELLxGENE data
portal provides a graphical user interface to collections of
single-cell sequence data processed in standard ways to 'count matrix'
summaries. The [cellxgenedp][] *R* / *Bioconductor* package provides
an *R*-based interface, allowing data discovery, viewing, and
downloading.

This vignette uses the [cellxgenedp][], [igraph][], and [visNetwork][]
packages. Install these with the following command

```{r install, eval = FALSE}
if (!"BiocManager" %in% rownames(installed.packages()))
    install.packages("BiocManager", repos = "https://cloud.r-project.org")
pkgs <- c("cellxgenedp", "igraph", "visNetwork")
required_pkgs <- pkgs[!pkgs %in% rownames(installed.packages())]
BiocManager::install(required_pkgs)
```

[cxgdp]: https://cellxgene.cziscience.com/
[igraph]: https://CRAN.R-project.org/package=igraph
[visNetwork]: https://CRAN.R-project.org/package=visNetwork
[cellxgenedp]: https://mtmorgan.github.io/cellxgenedp
[OLSr]: https://mtmorgan.github.io/OLSr

Start by loading the [OLSr][] and [cellxgenedp][] packages

```{r library-setup, message = FALSE}
library(OLSr)
library(cellxgenedp)
```

# CELLxGENE

A central element of the CELLxGENE resource is the collection of
[datasets][]. These are discovered programmatically using

```{r cellxgene-db}
db <- db()
datasets(db)
```

Data sets are classified using ontologies exposed by [cellxgenedp][] to
the user as 'facets'.

```{r cellxgene-facets}
facets <- facets(db)
facets
```

Some insight into the ontologies used to construct facets can be
gained by extracting the ontology from the `ontology_term_id`.

```{r cellxgene-ontology}
facets |>
    ## a small number of terms are not from any ontology; these do not
    ## have a ':' in the ontology_term_id
    filter(grepl(":", ontology_term_id, fixed = TRUE)) |> 
    ## for the remainder, discover the ontology from the ontology_term_id
    mutate(ontology = tolower(sub("(.*):.*", "\\1", ontology_term_id))) |>
    count(facet, ontology)
```

## Annotating the 'disease' facet

Our example will work with the 'disease' facet, based on the 'mondo'
ontology. Check that the ontology is in the OLS.

```{r mondo}
"mondo" %in% get_ontologies()$id
```

Retrieve all terms from the ontology.

```{r mondo-terms}
mondo <- get_terms("mondo")
mondo
```

There are `r NROW(mondo)` terms in this ontology; only a small subset
of `r NROW(facets(db, "disease"))` terms are used in CELLxGENE
datasets. Annotate the disease facet with information from the OLS.

```{r disease-terms}
disease_terms <-
    facets(db, "disease") |>
    ## join based on identity between CxG's 'ontology_term_id' and
    ## OLS's 'obo_id'
    left_join(mondo, by = c("label", ontology_term_id = "obo_id")) |>
    ## for consistency with OLSr convention, rename the
    ## 'ontology_term_id' to 'obo_id'
    rename(obo_id = "ontology_term_id")
disease_terms
```

Note that the 'normal' term with OBO id 'PATO:0000461' used by
CELLxGENE is not from the 'mondo' ontology; we could discover
information about this term by also retrieving the 'pato' ontology
from the OLS, or using `get_term("pato", "PATO:0000461")`.

[datasets]: https://cellxgene.cziscience.com/datasets

## Identifying carcinoma datasets

CELLxGENE presents the disease facet as a flat list of terms, although
the [datasets][] are presented with an *ad hoc* filter to group terms
and facilitate selection of datasets containing samples from multiple
terms. A more principled approach is to use the ontology to identify
terms of interest. 

Suppose one is interested in all datasets annotated with disease term
derived from 'carcinoma'; this term does not itself appear in the
disease facet. Find the `obo_id` of the 'carcinoma' term in the the
mondo ontology, and use that to find descendants.

```{r carcinoma-term-id}
carcinoma_id <-
    mondo |>
    filter(label == "carcinoma") |>
    pull("obo_id")
carcinoma <- get_descendants("mondo", carcinoma_id)
```

There are `r NROW(carcinoma)` terms descending from 'carcinoma'. Terms
used by the disease facet are

```{r carcinoma-disease-annotation}
disease_terms |>
    inner_join(carcinoma) |>
    select(obo_id, n, label, description)
```

The `facets_filter()` function of [cellxgenedp][] makes it easy to
identify datasets using terms derived from 'carcinoma', perhaps in
addition filtering to single type of assay.

```{r carcinoma-datasets}
datasets_of_interest <-
    datasets(db) |>
    filter(
        facets_filter(disease, "ontology_term_id", carcinoma$obo_id),
        facets_filter(assay, "label", "10x 3' v3")
    )
datasets_of_interest
```

There are `r NROW(datasets_of_interest)` datasets. Further steps, such
as visualization using CELLxGENE or downloading datasets for local
analysis, are described in the [cellxgenedp][] vignette.

# Ontologies as graphs

It can be useful to compute on ontologies as graphs. Use
`get_descendants_graph()` to retrieve, recursively, the children of
the 'carcinoma' term. This step can take one or two minutes, and
requires one API call per child. The API calls are memoised, so the
process is time consuming only the first time the graph is
constructed.

```{r get-descendants}
carcinoma_graph <-
    ## many calls to the OLS API; slow the first time
    get_descendants_graph("mondo", carcinoma_id, mondo)
```

`get_descendants_graph()` returns a `olsr_graph` object, which
contains information about the type of graph (ancestors or
descendants), the ontology used to construct the graph, and the ids
used to seed the query.

```{r carcinoma-graph}
carcinoma_graph
```

The object can be queried from relevant components, for instance
annotation of each node

```{r graph-nodes}
olsr_graph_nodes(carcinoma_graph)
```

## Graph manipulation

There are `r NROW(olsr_graph_nodes(carcinoma_graph))` nodes in this
graph, but only a dozen or so terms used in the CELLxGENE disease
facet. We would like to compute on the graph so that it contains only
the nodes involved in paths from the 'carcinoma' term to terms used by
the disease facet.

Start by coercing the `olsr_graph` to an object that can be used by
the [igraph][] package.

```{r olsr-to-igraph}
ig <- olsr_graph_as_igraph(carcinoma_graph)
```

`olsr_graph_as_igraph()` uses the nodes of `carcinoma_graph` as vertex
attributes, and adds information about the relationship, ontology, and
id(s) as a graph attribute. Note that the OLS field `obo_id` has been
translated to the igraph attribute `name`.

```{r graph_attr}
igraph::vertex_attr_names(ig)
igraph::graph_attr(ig, "olsr_graph") |>
    str()
```

Identifying a subgraph of `ig` corresponds (I think!)  to calculating
a minimal directed Steiner tree. An algorithm for this does not exist
in [igraph][] or other *R* packages, so we must develop our own. The
solution defined in the `directed_steiner_tree_approx()` function is
to find the shortest paths from the 'carcinoma' node to each term in
the disease facet, and then to create a subgraph containing just those
nodes.

```{r steiner-tree}
directed_steiner_tree_approx <-
    function(graph, from, to)
{
    ## only an approximation! Use the union of all nodes on all
    ## shortest paths from 'from' to each of 'to'

    ## find shortest paths from node 'from' to each 'to' node
    paths <- igraph::shortest_paths(ig, from, to)$vpath

    ## identify nodes on the shortest paths
    shortest_paths_nodes <-
        lapply(paths, names) |>
        unlist() |>
        unique()

    ## create an induced subgraph on the shortest-path nodes
    igraph::induced_subgraph(ig, shortest_paths_nodes)
}
```

This solution is not necessarily efficient, and does not guarantee a
minimal tree, but it is perhaps good enough for the size of data
encountered in a typical ontology.

To use the algorithm, identify the starting node and the relevant
nodes of the disease facet, and use these as inputs

```{r steiner-subgraph}
keep_ids <- c(
    carcinoma_id,
    intersect(disease_terms$obo_id, carcinoma$obo_id)
)
ig1 <- directed_steiner_tree_approx(ig, head(keep_ids, 1), tail(keep_ids, -1))
ig1
```

The induced graph returned by `directed_steiner_tree_approx()`
includes vertex and graph annotations from the input graph, and can be
coerced to a `olsr_graph` object.

```{r igraph-to-olsr}
igraph_as_olsr_graph(ig1)
```

## Graph visualization

There are `r length(igraph::V(ig1))` nodes in this subgraph, few
enough for convenient interactive visualization using the
[visNetwork][] package.

Use igraph commands to color nodes used in graph construction (the
'from' and 'to' nodes) differently from nodes required to represent
paths between nodes. Use the igraph 'name' attribute (OBO id) as the
'title' (i.e., mouse-over tool-tip). The 'label' attribute is used to
label nodes.

```{r igraph-color-nodes}
## colorspace::qualitative_hcl(2, "Set 2")
palette <- c("#ED90A4", "#00C1B2")
igraph::V(ig1)$color <- palette[ (igraph::V(ig1)$name %in% keep_ids) + 1L]
igraph::V(ig1)$title <- igraph::V(ig1)$name # obo_id
## V(ig1)$label             # 'label' (short description) from OLS
```

Create an interactive visualization using `visIgraph()`. The
'sugiyama' layout is appropriate for directed tree-like
graphs. `idToLabel = FALSE` ensures that the label vertex attribute,
rather than the internal igraph id, is used to label nodes.

Note that the graph is interactive -- zoom in and out, mouse-over and
select nodes, etc.

```{r visIgraph}
visNetwork::visIgraph(ig1, layout = "layout_with_sugiyama", idToLabel = FALSE)
```

A more flexible alternative for visualization is to translate an
`olsr_graph` to a `visNetwork` object, and to transform that for
visualization.

```{r visNetwork}
carcinoma_subgraph <- igraph_as_olsr_graph(ig1)
olsr_graph_as_visNetwork(carcinoma_subgraph) |>
    visNetwork::visEdges(arrows = "to") |>
    visNetwork::visHierarchicalLayout(
        sortMethod = "directed",
        direction = "LR"
    )
```

# Session information

```{r session-info}
sessionInfo()
```
