---
title: "1. Introduction to OLSr"
author:
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY, US
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{1. Introduction to OLSr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The [Ontology Lookup Service][OLS] ('OLS') is a repository of
biomedical ontologies provided by the European Bioinformatics
Institute (EMBL-EBI). [OLSr][] provides an interface to the OLS
service, allowing discovery of ontologies, terms, and
relations. Results are presented in a 'tidy' framework, so that
manipulation tasks are easily accomplished with verbs such as
'filter()', 'select()', and 'mutate()'. An effort is made to integrate
with other packages useful for more comprehensive biological and
ontological analysis.

[OLS]: https://www.ebi.ac.uk/ols4/

Written: 20 November, 2023

# Installation

Install [OLSr][] from GitHub using the [remotes][] package.

```{r install, eval = FALSE}
if (!"remotes" in rownames(installed.packages()))
    install.packages("remotes", repos = "https://cran.r-project.org")
remotes::install_github("mtmorgan/OLSr")
```

Attach the library to the current session. Most functions in [OLSr][]
return results as a `tibble`, so the [dplyr][] package is also
attached.

```{r library, message = FALSE}
library(OLSr)
```

[OLSr]: https://mtmorgan.github.io/OLSr
[remotes]: https://CRAN.R-project.org/package=remotes
[dplyr]: https://CRAN.R-project.org/package=dplyr

# Ontologies

Start by discovering ontologies available in the OLS.

```{r ontologies}
onto <- get_ontologies()
onto
```

This vignette uses for examples the ontology with title 'Cell
Ontology', used by the [CELLxGENE][] project to classify cell
types. Find the record corresponding to this ontology, using
`glimpse()` to provide a convenient view of the ontology description.

[CELLxGENE]: https://cellxgene.cziscience.com

```{r cell-ontology}
onto |>
    filter(title == "Cell Ontology" ) |>
    glimpse()
```

Important information includes the 'id' of the ontology (used in many
subsequent steps), the ontology home page, and the number of terms in
the ontology.

```{r, cell-ontology-2}
onto |>
    filter(title == "Cell Ontology") |>
    select(id, homepage, numberOfTerms)
```

Information on individual ontologies is also available with
`get_ontology()`; this record includes a link to the ontology page in
the OLS. The home page is a good starting place for understanding the
ontology in a more visual and interactive environment.

```{r get-ontology}
cl <- get_ontology("cl")
glimpse(cl)

cl |> pull("_links")
```

# Terms

Terms are central to using an ontology. Retrieve all terms from the
'cl' ontology. By default, only terms defined in the ontology *per se*
are used; add the argument `all_ontologies = TRUE` to this and
subsequent functions to also include terms referenced in other
ontologies. This query can take several minutes to complete when first
executed, but the result is cached locally (as discussed in greater
detail below) so subsequent use is quick.

```{r get-terms}
terms <- get_terms("cl")
terms
```

Glimpse the second term to get a sense of available information.

```{r term-slice}
terms |>
    slice(1) |>
    glimpse()
```

## Hierarchical data

Some columns in the `terms` table represent hieararchical (relational,
1:many mapping) data, e.g., the `in_subset` column

```{r hierarchy}
terms |>
    select(label, obo_id, in_subset)
```

Individual terms apparently belong to 0 or more subsets.

```{r in_subset-count}
terms |>
    count(lengths(in_subset))
```

The R for Data Science chapter on [hierarchical data][] suggests using
[tidyr][] functions `unnest_wider()` and `unnest_longer()` to expand
such data.

[tidyr]: https://CRAN.R-project.org/package=tidyr

```{r in_subset-unnest-longer}
terms |>
    select(in_subset, obo_id, label) |>
    tidyr::unnest_longer(in_subset)
```

Defined subsets and the number of terms in each can be computed as

```{r}
terms |>
    tidyr::unnest_longer(in_subset) |>
    count(in_subset, sort = TRUE)
```

Use `filter()` with a helper function that summarizes subsets of
individual terms to identify the `human_reference_atlas` subset. Using
`vapply()` (rather than, e.g., `sapply()`) can be considered a best
practice, so that the filter function is guaranteed to return a scalar
logical value regardless of input length.

```{r human_reference_atlas-subset}
is_human_reference_atlas <- function(x)
    vapply(x, \(x) any(x == "human_reference_atlas"), logical(1))

human_reference_atlas <-
    terms |>
    filter(is_human_reference_atlas(in_subset))

human_reference_atlas
```

There are `r NROW(human_reference_atlas)` terms in the
`human_reference_atlas` subset.

[hierarchical data]: https://r4ds.hadley.nz/rectangling

## Individual terms

Suppose we are interested in B cells. Discover term labels starting
with `"B cell"`, selecting useful fields for display. For subsequent
work, create a variable capturing the `obo_id` of the `"B cell"` term.

```{r B-cell-terms}
terms |>
    filter(startsWith(label, "B cell")) |>
    select(label, obo_id, description)
b_cell <- # "CL:0000236"
    terms |>
    filter(label == "B cell") |>
    pull(obo_id)
```

Slightly richer information on individual terms is available with
`get_term()`, using the ontology `id` and term `obo_id` as keys.

```{r B-cell-term}
get_term("cl", b_cell) |>
    glimpse()
```

# Relatives

The functions `get_parents()` and `get_children()` retrieve parent and
child terms of individual terms; `get_ancestors()` and
`get_descendants()` return all ancestors (to the root of the ontology)
or descendants (where the final term repreents a leaf). For B cells
the ancestors are

```{r b-cell-ancestors}
b_cell_ancestors <-
    get_ancestors("cl", b_cell)
```


Ancestors and descendants also contain the `in_subset` column, so
these groups are easily filtered to a particlur subset, e.g.,

```{r b-cell-descendants-human-reference-atlas}
b_cell_ancestors |>
    filter(is_human_reference_atlas(in_subset))
```

Terms without children or descendants return a tibble without any
rows.

```{r}
get_children("cl", "CL:0002350")
get_descendants("cl", "CL:0002350")
```

# Cache management

OLSr uses an on-disk cache to improve performance and
reliability. Each query to the OLS server is cached on first use, so
that subsequent responses can be read from the local disk. This
greatly increases performance, and allows use of previously retrieved
resources when off-line.

OLSr uses the [memoise][] and [cachem][] packages to implement the
cache. The disk cache is implemented using cachem's `cache_disk()`
function using the default cache size (1 Gb) and `"lsu"` ('least
recently used') eviction policy. Objects remain in the cache for 30
days.

[memoise]: https://CRAN.R-project.org/package=memoise
[cachem]: https://CRAN.R-project.org/package=cachem

A risk of using a cache is that the cache contains an 'old' version
of the response. Circumvent this risk by manually manipulating the
cache, e.g., removing objects that a known to be too old.

Use `cache_directory()` to discover the location of the cache.

```{r cache-directory}
cache_directory()
```

This location follows conventions of `tools::R_user_dir()`. All files
in this directory have been created by [OLSr][].

`cache_info()` is a convenience function to summarize files in the
cache.

```{r cache-info}
cache_info()
```

Each file represents the response to a single memoised function
call. The file name is a hash of the function definition and
arguments; it is not possible to associate individual files with
specific function calls. Nonetheless, it can be informative to
identify files modified more than, say 1 day ago, or the most recently
modified file, or perhaps the total size of files in the cache.

```{r cache-info-uses}
## cache objects modified in the last 24 hours
cache_info() |>
    filter(mtime > Sys.time() - 24 * 60 * 60)

## most recently modified file
cache_info() |>
    arrange(desc(mtime)) |>
    slice(1)

## total cache size
cache_info() |>
    summarize(size = sum(size)) |>
    pull(size) |>
    structure(class = "object_size") |>
    format(units = "auto")
```

In special cases it may be useful to invalidate the cache using
`unlink()` on the `cache_directory()`, or on invalidate particular
records using `unlink()` on file paths created using `file.path()`,
`cache_directory()` and individual `file` names from `cache_info()`.

# Session information {.unnumbered}

```{r}
sessionInfo()
```
